---
title: "intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```


# Introduction

This vignette of package **`fmx`** ([CRAN](https://CRAN.R-project.org/package=fmx), [Github](https://github.com/tingtingzhan/fmx)) documents ..

## Note to Users

Examples in this vignette require that the `search` path has

```{r setup}
library(fmx)
library(ggplot2)
```


## Terms and Abbreviations

```{r echo = FALSE, results = 'asis'}
c(
  '', 'Forward pipe operator', '`?base::pipeOp` introduced in `R` 4.1.0', 
  '`CRAN`, `R`', 'The Comprehensive R Archive Network', 'https://cran.r-project.org',
  '`curve`', 'Function plots', '`graphics::curve`',
  '`mc.cores`', 'Number of CPU cores to use', '`parallel::mclapply`; `parallel::detectCores`',
  '`object.size`', 'Memory allocation', '`utils::object.size`',
  '`save`, `xz`', 'Save with `xz` compression', '`base::save(., compress = \'xz\')`; `base::saveRDS(., compress = \'xz\')`; https://en.wikipedia.org/wiki/XZ_Utils', 
  '`S3`, `generic`, `methods`', '`S3` object oriented system',  '`base::UseMethod`; `utils::methods`; `utils::getS3method`; https://adv-r.hadley.nz/s3.html',
  '`S4`, `generic`, `methods`', '`S4` object oriented system',  '`base::isS4`; `methods::setClass`; `methods::getMethod`; https://adv-r.hadley.nz/s4.html',
  '`search`', 'Search path', '`base::search`'
) |>
  matrix(nrow = 3L, dimnames = list(c('Term / Abbreviation', 'Description', 'Reference'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable(format = 'html') 
# ?knitr::kable
# default: `|` shown as &...
# format = 'html': `>` shown as &..
```


# `S4` Class `'fmx'`

## Definition

The `S4` class stores the parameters of a mixture of distribution. Currently the `norm`al distribution and Tukey's `GH` distribution is supported.

Example below shows a mixture of two normal distributions $N(0,1)$ and $N(3,1.3)$ with $(50\%, 50\%)$ mixing probability (i.e., the mixing probability $w$ is normalize to so that the summation of $w$ is 1).
```{r}
(e1 = fmx('norm', mean = c(0,3), sd = c(1,1.3), w = c(1, 1)))
class(e1)
isS4(e1) # TRUE
```

Example below shows a mixture of two Tukey's $g$-and-$h$ distributions $(A,B,g,h)=(0,1,.2,.2)$ and $(3,1,.3,.1)$ (i.e., by default the location parameter $A=0$ and the scale parameter $B=1$) with $(20\%,30\%)$ mixing probability.

```{r}
(e2 = fmx('GH', A = c(0,3), g = c(.2, .3), h = c(.2, .1), w = c(2, 3)))
```

The `fmx` framework also accepts a single-component distribution.

```{r}
(e3 = fmx('GH', g = .2, h = .2))
```


## Density

```{r}
curve(dfmx(x, dist = e1), xlim = c(-3,7), n = 501L, ylab = 'Density')
```

```{r fig.height=2}
ggplot() + geom_function(fun = dfmx, args = list(dist = e1), n = 501L) + xlim(-3,7)
```

## Probability

```{r}
curve(pfmx(x, dist = e1), xlim = c(-3,7), n = 501L, ylab = 'Probability')
```

```{r fig.height=2}
ggplot() + geom_function(fun = pfmx, args = list(dist = e1), n = 501L) + xlim(-3,7)
```

## Quantile

```{r}
curve(qfmx(x, dist = e1), xlim = c(0,1), n = 501L, ylab = 'Quantile')
```

```{r fig.height=2}
ggplot() + geom_function(fun = qfmx, args = list(dist = e1), n = 501L) + xlim(0,1)
```

## Simulation

```{r}
e1 |> rfmx(n = 1e3L) |> hist(main = '1000 obs from e1')
```

## Visualize Multiple `'fmx'` Objects

```{r fig.width=5}
ggplot() + 
 geom_function(fun = qfmx, args = list(dist=e1), mapping = aes(color='Normal'), n = 501L) + 
 geom_function(fun = qfmx, args = list(dist=e2), mapping = aes(color='Tukey gh'), n = 501L) + 
 scale_x_continuous(labels = scales::percent, limits = c(.01, .99)) +
 labs(x = NULL, y = NULL, color = 'Mixture\nModels')
```



```{r}
# to use [rfmx] without \pkg{fmx}
(d = fmx(distname = 'GH', A = c(-1,1), B = c(.9,1.1), g = c(.3,-.2), h = c(.1,.05), w = c(2,3)))
d@pars
set.seed(14123); x = rfmx(n = 1e3L, dist = d)
set.seed(14123); x_raw = rfmx(n = 1e3L,
 distname = 'GH', K = 2L,
 pars = rbind(
  c(A = -1, B = .9, g = .3, h = .1),
  c(A = 1, B = 1.1, g = -.2, h = .05)
 ), 
 w = c(.4, .6)
)
stopifnot(identical(x, x_raw))
```



```{r}
(d0 = fmx('GH', A = c(1,4), g = c(.2,.1), h = c(.05,.1), w = c(1,1)))
(c0 = fmx_constraint(d0))
user_constraint(character(), distname = 'GH', K = 2L) # equivalent

(d1 = fmx('GH', A = c(1,4), g = c(.2,0), h = c(0,.1), w = c(1,1)))
(c1 = fmx_constraint(d1))
user_constraint(c('g2', 'h1'), distname = 'GH', K = 2L) # equivalent

(d2 = fmx('GH', A = c(1,4), g = c(.2,0), h = c(.15,.1), w = c(1,1)))
(c2 = fmx_constraint(d2))
user_constraint('g2', distname = 'GH', K = 2L) # equivalent
```


```{r}
(d0 = fmx('GH', A = c(1,4), g = c(.2,.1), h = c(.05,.1), w = c(1,1)))
(c0 = fmx_constraint(d0))
user_constraint(distname = 'GH', K = 2L, x = character()) # equivalent

(d1 = fmx('GH', A = c(1,4), g = c(.2,0), h = c(0,.1), w = c(1,1)))
(c1 = fmx_constraint(d1))
user_constraint(distname = 'GH', K = 2L, x = c('g2', 'h1')) # equivalent

(d2 = fmx('GH', A = c(1,4), g = c(.2,0), h = c(.15,.1), w = c(1,1)))
(c2 = fmx_constraint(d2))
user_constraint(distname = 'GH', K = 2L, x = 'g2') # equivalent
```


```{r}
(d0 = fmx('GH', A = c(1,4), g = c(.2,.1), h = c(.05,.1), w = c(1,1)))
getTeX(d0)

(d1 = fmx('GH', A = c(1,4), g = c(.2,0), h = c(0,.1), w = c(1,1)))
getTeX(d1)

(d2 = fmx('GH', A = c(1,4), g = c(.2,0), h = c(.15,.1), w = c(1,1)))
getTeX(d2)
```


```{r}
(d = fmx('norm', mean = c(1, 4, 7), w = c(1, 1, 1)))
d[1:2]
```





```{r}
(d2 = fmx('GH', A = c(1,6), B = 2, g = c(0,.3), h = c(.2,0), w = c(1,2)))
moment_fmx(d2)
```



```{r}
m = c(-1.5, 1.5)
s = c(.9, 1.1)
sk = c(.2, -.3)
kt = c(.5, .75)
w = c(2, 3)
(d1 = moment2fmx(distname='GH', w=w, mean=m, sd=s, skewness=sk, kurtosis=kt))
moment_fmx(d1)
(d2 = moment2fmx(distname='st', w=w, mean=m, sd=s, skewness=sk, kurtosis=kt))
moment_fmx(d2)
library(ggplot2)
ggplot() + 
 geom_function(aes(color = 'GH'), fun = dfmx, args = list(dist=d1), n = 1001) + 
 geom_function(aes(color = 'st'), fun = dfmx, args = list(dist=d1), n = 1001) +
 xlim(-5, 6)
# two curves looks really close, but actually not identical
x = rfmx(n = 1e3L, dist = d1)
range(dfmx(x, dist = d1) - dfmx(x, dist = d2))
```






# Convert to `'fmx'` Class

## ... from `fitdistrplus::fitdist` object
```{r}
library(fitdistrplus)
# ?fitdist
data(endosulfan, package = 'fitdistrplus')
ATV <- subset(endosulfan, group == 'NonArthroInvert')$ATV
log10ATV <- log10(ATV)
fln <- fitdist(log10ATV, distr = 'norm')
(fln2 <- as.fmx(fln))
hist.default(log10ATV, freq = FALSE)
curve(dfmx(x, dist = fln2), xlim = range(log10ATV), add = TRUE)
```


## ... from objects created by package **`mixtools`**
```{r}
library(mixtools)
(wait = as.fmx(normalmixEM(faithful$waiting, k = 2)))
hist.default(faithful$waiting, freq = FALSE)
curve(dfmx(x, dist = wait), xlim = range(faithful$waiting), add = TRUE)
```

## ... from objects created by package **`mixsmsn`** 
```{r}
library(mixsmsn)
# ?smsn.mix
arg1 = c(mu = 5, sigma2 = 9, lambda = 5, nu = 5)
arg2 = c(mu = 20, sigma2 = 16, lambda = -3, nu = 5)
arg3 = c(mu = 35, sigma2 = 9, lambda = -6, nu = 5)
set.seed(120); x = rmix(n = 1e3L, p=c(.5, .2, .3), family = 'Skew.t', 
  arg = list(unname(arg1), unname(arg2), unname(arg3)))
```

```{r}
# Skew Normal
class(m1 <- smsn.mix(x, nu = 3, g = 3, family = 'Skew.normal', calc.im = FALSE))
mix.hist(y = x, model = m1)
m1a = as.fmx(m1, data = x)
(l1a = logLik(m1a))
hist(x, freq = FALSE)
curve(dfmx(x, dist = m1a), xlim = range(x), add = TRUE)
```

```{r}
library(mixsmsn)
# ?smsn.mix
arg1 = c(mu = 5, sigma2 = 9, lambda = 5, nu = 5)
arg2 = c(mu = 20, sigma2 = 16, lambda = -3, nu = 5)
arg3 = c(mu = 35, sigma2 = 9, lambda = -6, nu = 5)
set.seed(120); x = rmix(n = 1e3L, p=c(.5, .2, .3), family = 'Skew.t', 
  arg = list(unname(arg1), unname(arg2), unname(arg3)))

# Normal
class(m2 <- smsn.mix(x, nu = 3, g = 3, family = 'Normal', calc.im = FALSE))
mix.hist(y = x, model = m2)
m2a = as.fmx(m2, data = x)
hist(x, freq = FALSE)
curve(dfmx(x, dist = m2a), xlim = range(x), add = TRUE)
```


```{r eval=FALSE}
# mixsmsn::smsn.mix with option `family = 'Skew.t'` is slow

library(mixsmsn)
# ?smsn.mix
arg1 = c(mu = 5, sigma2 = 9, lambda = 5, nu = 5)
arg2 = c(mu = 20, sigma2 = 16, lambda = -3, nu = 5)
arg3 = c(mu = 35, sigma2 = 9, lambda = -6, nu = 5)
set.seed(120); x = rmix(n = 1e3L, p=c(.5, .2, .3), family = 'Skew.t', 
  arg = list(unname(arg1), unname(arg2), unname(arg3)))

# Skew t
class(m3 <- smsn.mix(x, nu = 3, g = 3, family = 'Skew.t', calc.im = FALSE))
mix.hist(y = x, model = m3)
m3a = as.fmx(m3, data = x)
hist(x, freq = FALSE)
curve(dfmx(x, dist = m3a), xlim = range(x), add = TRUE)
(l3a = logLik(m3a))
stopifnot(all.equal.numeric(AIC(l3a), m3$aic), all.equal.numeric(BIC(l3a), m3$bic))
```

```{r}
library(mixsmsn)
# ?smsn.mix
arg1 = c(mu = 5, sigma2 = 9, lambda = 5, nu = 5)
arg2 = c(mu = 20, sigma2 = 16, lambda = -3, nu = 5)
arg3 = c(mu = 35, sigma2 = 9, lambda = -6, nu = 5)
set.seed(120); x = rmix(n = 1e3L, p=c(.5, .2, .3), family = 'Skew.t', 
  arg = list(unname(arg1), unname(arg2), unname(arg3)))

# t
class(m4 <- smsn.mix(x, nu = 3, g = 3, family = 't', calc.im = FALSE))
mix.hist(y = x, model = m4)
# as.fmx(m4, data = x) # not ready yet!!
```


